---
title: "A light introduction to runoutSIM"
author: "Jason Goetz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A light introduction to runoutSIM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `runoutSIM` package is developed for regional-scale runout simulations using random walks. This vignette provides an introduction to applying `runoutSIM` for regional debris flow simulation using a subcatchment of the upper Maipo river basin, Chile as an example. In this vignette, we will:

-   Load and visualize spatial data
-   Simulate runout from single and multiple source cells
-   Visualize simulation results as rasters and interactive maps

# Loading packages and reading spatial data

Spatial handling in the `runoutSIM` package is supported by the `sf` (vector data) and `terra` (gridded/raster data) packages. In this example, we load/read a 12.5 m spatial resolution DEM, debris flow runout polygons, and corresponding source points mapped for each polygon. We also create a hillshade model to help with visualization of the results.

**Tip:** Currently, this package is designed to work with data in a UTM coordinate reference system (CRS). You will need to project your own data to a local UTM CRS for it to work.

```{r, fig.show='hold'}
# load packages
library(runoutSim)
library(terra)
library(sf)

# Load digital elevation model (DEM)
dem <- rast("C:/GitProjects/runoutSIM/Dev/Data/elev_fillsinks_WangLiu.tif")

# Round elevation to nearest meter [source was integer]
dem <- round(dem)

# Compute hillshade for visualization 
slope <- terrain(dem, "slope", unit="radians")
aspect <- terrain(dem, "aspect", unit="radians")
hill <- round(shade(slope, aspect, 40, 270, normalize = TRUE))

# Load debris flow runout source points and polygons
source_points <- st_read("C:/GitProjects/runoutSIM/Dev/Data/debris_flow_source_points.shp")
runout_polygons <- st_read("C:/GitProjects/runoutSIM/Dev/Data/debris_flow_runout_polygons.shp")

# Load basin boundary
bnd_catchment <- st_read("C:/GitProjects/runoutSIM/Dev/Data/basin_rio_olivares.shp")

```

# Creating interactive maps

Using the `leafmap()` function, we can quickly create an interactive leaflet map for viewing the data. Similar to `dplyr`, we use the `%>%` operator to pipe together multiple `leafmap()` calls to create a combined leaflet map with multiple layers.

```{r, fig.show='hold', out.width="100%", out.height="500px"}
map <- leafmap(bnd_catchment, color = '#f7f9f9', fill_color = '#FF000000', weight = 4) %>%
    leafmap(runout_polygons) %>%
    leafmap(source_points, color = '#e74c3c') %>%
    leafmap(hill, palette = grey(150:255/255), opacity = 1, add_legend = FALSE) %>%
    leafmap(dem, palette = viridis::mako(10), opacity = 0.6)
map
```

# Simulate runout from a source cell

Runout paths are simulated from individual source cells, defined as x, y coordinates. For example, if we are simulating runout from a single source cell defined by a source point (vector data), we use `sf::st_coordinates()` to extract the x, y values.

```{r}
# Select a single debris flow and source point for the example
runout_polygon <- runout_polygons[31,]

# Get corresponding source point
source_point  <- st_filter(st_as_sf(source_points), st_as_sf(runout_polygon))

# Get coordinates of source point
source_crds <- st_coordinates(source_point)
print(source_crds)
```

Now that we have the coordinates, we can run the random walk runout simulations using `runoutSim()`. Where:

-   `mu` and `md` are the sliding friction coefficient and mass-to-drag parameters of the two-parameter friction model controlling runout distance

-   `slp_thresh`, `exp_div`, and `per_fct` are parameters controlling the dispersion of the random walks

-   `walks` is the number of random walk iterations

```{r, fig.show='hold'}
sim_paths = runoutSim(dem = dem, 
                      xy = source_crds, 
                      mu = 0.08, 
                      md = 140, 
                      slp_thresh = 35, 
                      exp_div = 2.5, 
                      per_fct = 1.95, 
                      walks = 1000)

# plot structure of sim_paths
str(sim_paths)
```

`runoutSim()` returns a list containing:

-   The cell number of the DEM indicating the location of the source cell

-   Traverse frequencies for each grid cell

-   Maximum velocities (m/s) for each grid cell

These values are stored as cell numbers (e.g., `$all_indices`). Although not calculated here, `prob_connect` provides the probability the source cell connected to a feature of interest.

## Converting results to a raster

To visualize the results, we convert the output from `runoutSim()` to a raster.

```{r, fig.height=5, fig.width = 7, fig.align='center'}
## Convert paths to raster with cell transition frequencies
paths_raster <- walksToRaster(sim_paths, dem, method = "freq")

# Plot results
paths_raster <- crop(paths_raster, ext(runout_polygon)+500)
plot(crop(hill, ext(runout_polygon)+500), col=grey(150:255/255), legend=FALSE,
     mar=c(2,2,1,4), main = "Runout traverse frequency")
plot(paths_raster, legend = T, alpha = 0.5, add = TRUE)
plot(st_geometry(runout_polygon), add = TRUE)
plot(source_point, add = TRUE)
```

In `walksToRaster()`, we use `method = "freq"` to return traverse frequencies. We can also return traverse (ECDF) probabilities by using `method = "cdf_prob"`.

We use `velocityToRaster()` to get a raster of velocity values (m/s).

```{r, fig.height=5, fig.width = 7, fig.align='center'}
## Convert paths to raster with max. velocities
vel_raster <- velocityToRaster(sim_paths, dem)

# Plot results
vel_raster <- crop(vel_raster, ext(runout_polygon)+500)
plot(crop(hill, ext(runout_polygon)+500), col=grey(150:255/255), legend=FALSE,
     mar=c(2,2,1,4), main = "Runout velocity")
plot(vel_raster, col = map.pal('plasma'), legend = TRUE, alpha = 0.5, add = TRUE)
plot(st_geometry(runout_polygon), add = TRUE)
plot(source_point, add = TRUE)
```

# Simulate runout from multiple-sources

To simulate runout from multiple source cell locations, we run `runoutSim()` iteratively, changing the xy coordinates during each iteration.

To create a list of xy coordinates from source points (vector), we use the `makeSourceList()` function.

```{r, fig.show='hold', out.width="100%", out.height="500px"}
# Get coordinates of source points to create a source list object
source_l <- makeSourceList(source_points)

# Perform random walk simulation for each source point
sim_runs <- list()

for(i in 1:length(source_l)){
  
  sim_runs[[i]] <- runoutSim(dem = dem, xy = source_l[[i]], 
                              mu = 0.08, 
                              md = 40, 
                              slp_thresh = 40, 
                              exp_div = 3, 
                              per_fct = 1.9, 
                              walks = 1000)
}



```

We can now convert this list of multiple random walks from different source cells to raster data using `walksToRaster()` and `velocityToRaster()`.

```{r, fig.show='hold', fig.height=4, fig.width = 7, fig.align='center'}
# Coerce results to a raster
trav_freq <- walksToRaster(sim_runs, method = "freq", dem)
vel_ms <- velocityToRaster(sim_runs, dem, method = "max")
trav_prob <- walksToRaster(sim_runs, method = "max_cdf_prob", dem)


# Plot random walks from mulitple source cells
par(mfrow = c(1,3))
plot(hill, col=grey(150:255/255), legend=FALSE, mar=c(2,2,1,4), 
     main = "Traverse frequency")
plot(trav_freq, add = TRUE, alpha = 0.7)

plot(hill, col=grey(150:255/255), legend=FALSE, mar=c(2,2,1,4), 
     main = "Traverse probability (max.)")
plot(trav_prob, add = TRUE, alpha = 0.7)

plot(hill, col=grey(150:255/255), legend=FALSE, mar=c(2,2,1,4), 
     main = "Velocity (m/s)")
plot(vel_ms, col = map.pal("plasma"), add = TRUE, alpha = 0.7)


```

We can also view the results through an interactive leaflet map.

```{r, fig.show='hold', out.width="100%", out.height="500px"}

# Create interactive leaflet map
sim_map <- 
  # Debris flow observations
  leafmap(runout_polygons, opacity = 0.3, label = "Runout polygons") %>%
  leafmap(source_points, color = '#e74c3c', label = "Source points") %>%
  # Modelling results
  leafmap(trav_prob,  palette = viridis::viridis(10, direction = -1), 
          label = "Traverse prob") %>%
  leafmap(vel_ms,  palette = viridis::plasma(10, direction = -1),
          label = "Velocity")

# Start with mapped zoomed in
sim_map <- leaflet::setView(sim_map, lng = -70.13694, lat = -33.3703, zoom = 13)


# Hide layers
sim_map <- leaflet::hideGroup(sim_map, c("Velocity", "Source points"))

sim_map

```
